Inferring the distribution and rate of Earth-like exoplanets \\ from noisy individual-planet inferences

### 2014-05-14 document/ms.tex
removing appendix

A very simple example is the familiar procedure of making a histogram.
If you model the function $p(\{\entry_k\}\,|\,\ratepars)$ as a piecewise
constant rate function---where the bin heights are the parameters---and if the
uncertainties on the catalog are negligible compared to the bin widths then
the maximum marginalized likelihood solution for \ratepars\ is a histogram of
the catalog entries (see \app{histogram} for the derivation).
The case of non-negligible uncertainties is described by \citet{hogge} using a
method similar to the one discussed here.

@@

For each catalog, we directly apply the inverse-detection-efficiency
procedure as implemented by \citealt{petigura}\footnote{Our implementation
exactly reproduces their results when applied to the published catalog.} and
our probabilistic method, marginalizing over the hyperparameters of the
Gaussian process regularization.
\Fig{smooth-results} and \fig{simulation-results} show the results of this
analysis in both cases.
In particular, the side panels compare the marginalized occurence rate
functions in period and radius to the true functions that were used to
generate the catalogs.
\Fig{smooth-results} shows that even if the true occurence rate is a smooth
function, the rate inferred by the inverse-detection-efficiency method can
appear to have sharp features.
In this first example---where the true distribution is well described by our
Gaussian process model---the probabilistic inference of the occurence rate is
both more precise and accurate.

@@

\acknowledgments
It is a pleasure to thank
    Erik Petigura (Berkeley),
    \etc
for helpful contributions to the ideas and code presented here.
This project was partially supported by the NSF (grant AST-0908357), and NASA
(grant NNX08AJ48G).
This research made use of the NASA \project{Astrophysics Data System}.

\appendix

\section{Maximum likelihood estimators for noiseless catalogs}
\sectlabel{histogram}

Equation XXXX gives the likelihood of a catalog $\catalog =
\{\entry_k\}_{k=1}^K$, conditioned on a particular choice of occurence rate
parameters \ratepars:
\begin{eqnarray}
\ln p(\catalog\,|\,\ratepars) &=& \sum_{k=1}^K \ln \rate(\entry_k)
    - \int \rate(\entry) \dd\entry \quad.
\end{eqnarray}
If we model the occurence rate as a piecewise constant function
\begin{eqnarray}
\rate (\entry) &=& \left \{\begin{array}{ll}
\ratepar_1 & \mathrm{if\,} \entry \in \binarea_1, \\
\cdots & \cdots \\
\ratepar_N & \mathrm{if\,} \entry \in \binarea_N, \\
0 & \mathrm{otherwise}
\end{array}\right.
\end{eqnarray}
where $\binarea_n$ is the volume of the bin $n$, the log likelihood becomes
\begin{eqnarray}
\ln p(\catalog\,|\,\ratepars) &=&
    \sum_{k=1}^K \sum_{n=1}^N \mathbf{1}[\entry_k \in \binarea_n]
        \ln\ratepar_n
    - \sum_{n=1}^N \ratepar_n\,\binarea_n \quad.
\end{eqnarray}
The maximum likelihood solution to this can be computed analytically
\begin{eqnarray}
\frac{\dd \ln p(\catalog\,|\,\ratepars)}{\dd \ratepar_m} &=&
    \frac{N_m}{\ratepar_m} - \binarea_m \\
\to \quad \ratepar_m &=& \frac{N_m}{\binarea_m}
\end{eqnarray}
where $N_m$ is the number of objects in the bin $m$.
This result is simply a \emph{normalized histogram} of the points.

Including the completeness function $\completeness(\entry)$ is as simple as
modeling the observable occurence rate as
\begin{eqnarray}
\rate (\entry) &=& \completeness(\entry) \, \left \{\begin{array}{ll}
\ratepar_1 & \mathrm{if\,} \entry \in \binarea_1, \\
\cdots & \cdots \\
\ratepar_N & \mathrm{if\,} \entry \in \binarea_N, \\
0 & \mathrm{otherwise}
\end{array}\right. \quad.
\end{eqnarray}
This leads to a similar analytic maximum likelihood solution
\begin{eqnarray}
\frac{\dd \ln p(\catalog\,|\,\ratepars)}{\dd \ratepar_m} &=&
    \frac{N_m}{\ratepar_m}
    - \int_{\binarea_m} \completeness(\entry)\dd\entry \\
\to \quad \ratepar_m &=& N_m \,
    \left [\int_{\binarea_m} \completeness(\entry)\dd\entry \right ]^{-1}
\end{eqnarray}
where the remaining integral is over the bin $m$.
This result is \emph{almost} the same as the inverse-detection-efficiency
procedure but each point is weighted by the bin-averaged efficiency instead of
the per-object efficiency.

>>>>

A very simple example is the familiar procedure of making a histogram.
If you model the function $p(\{\entry_k\}\,|\,\ratepars)$ as a piecewise
constant rate function---where the bin heights are the parameters---and if the
uncertainties on the catalog are negligible compared to the bin widths then
the maximum marginalized likelihood solution for \ratepars\ is a histogram of
the catalog entries.
The case of non-negligible uncertainties is described by \citet{hogge} using a
method similar to the one discussed here.

@@

For each catalog, we directly apply the inverse-detection-efficiency procedure
as implemented by \citealt{petigura}\footnote{Our implementation reproduces
their results when applied to the published catalog.} and our probabilistic
method, marginalizing over the hyperparameters of the Gaussian process
regularization.
\Fig{smooth-results} and \fig{simulation-results} show the results of this
analysis in both cases.
In particular, the side panels compare the marginalized occurence rate
functions in period and radius to the true functions that were used to
generate the catalogs.
\Fig{smooth-results} shows that even if the true occurence rate is a smooth
function, the rate inferred by the inverse-detection-efficiency method can
appear to have sharp features.
In this first example---where the true distribution is well described by our
Gaussian process model---the probabilistic inference of the occurence rate is
both more precise and accurate.

@@

\acknowledgments
We would like to thank Erik Petigura (Berkeley) for freely sharing his data
and code.
It is a pleasure to thank
\ldots
for helpful contributions to the ideas and code presented here.
This project was partially supported by the NSF (grant AST-0908357), and NASA
(grant NNX08AJ48G).
This research builds on collaborations between astronomers and statisticians
forged during a three week workshop on ``Modern Statistical and Computational
Methods for Analysis of Kepler Data'' at SAMSI in June 2013.
This research made use of the NASA \project{Astrophysics Data System}.
